
```{r}
renv::activate()
```

```{r}
library(dplyr)
```

#Filter's preserve argument
Filter preserve - 
decide whether to maintain groups that don't meet filtering conditions.
What happens when a group has no items?
It remains for future summarisation/other operations as NULL/NA.
So why do that?
If writing a report or whatnot, you may want the empty group to persist in the summary  table.
```{r}
mtcars %>%
  ggplot2::ggplot(ggplot2::aes(x= cyl, y = hp)) +
  ggplot2::geom_point()
```

Compare:
```{r}
mtcars %>%
  group_by(cyl) %>%
  filter(hp > 200) %>%
  summarise(x = mean(disp))
```


```{r}
mtcars %>%
  group_by(cyl) %>%
  filter(hp > 200, .preserve = TRUE) %>%
  summarise(x = mean(disp))

tibble::tibble(mtcars) %>% # come back to this
  filter(hp > 50, .by = cyl) %>%
  summarise(x = mean(disp))

tibble::tibble(mtcars) %>%
  filter(hp > mean(hp), .by = cyl)
```

```{r}
tibble::tibble(
  username = c("michael", "philippa", NA),
  source = c("Twitter", "Twitter", "Reddit"),
  text = c("not interesting", "interesting", "interesting")
) %>%
  mutate(username = ifelse(is.na(username), "na_xjd", username)) %>%
  filter(username != "michael")
```

```{r}
#Slice family
`slice()`
`slice_head()`
`slice_tail()`
`slice_min()`
`slice_max()`
`slice_sample()`
```

```{r}
mtcars %>%
  filter(row_number() == 10)
```

```{r}
mtcars %>%
  dplyr::slice(-1L)

mtcars %>%
  tibble::tibble() %>%
  dplyr::mutate(id = row_number(), .before = 1) %>%
  dplyr::slice(-1L, -2L, -3L)
```


Use formals to get the args, then use lapply to get the lengths
```{r}
slice_args <- list(
  slice = formals(dplyr::slice),
  slice_head = formals(dplyr::slice_head),
  slice_tail = formals(dplyr::slice_tail),
  slice_max = formals(dplyr::slice_max),
  slice_min = formals(dplyr::slice_min),
  slice_sample = formals(dplyr::slice_sample)
)

lapply(slice_args, length)

slice_args$slice_head
slice_args$slice
formals(dplyr::filter)
```


```{r}
?slice
?slice_head
?slice_tail
?slice_max
?slice_min
?slice_sample
```

```{r}
mtcars <- tibble::tibble(mtcars)
mtcars %>%
  dplyr::slice_max(prop = 0.25, 
                   order_by = hp,
                   # with_ties = FALSE
                   )

mtcars %>%
  dplyr::slice_min(
    prop = 0.25,
    # n = 2,
    order_by = hp,
    with_ties = FALSE)

mtcars %>%
  dplyr::group_by(cyl) %>%
  dplyr::slice_max(prop = 0.25, order_by = hp)

mtcars %>%
  dplyr::slice_head(n = 5)
```



```{r}
?dplyr::slice_sample

mtcars %>%
  dplyr::mutate(weight = 
    dplyr::case_when(cyl == 4 ~ 0.2,
                     cyl == 6 ~ 0.4,
                     cyl == 8 ~ 0.4)
  )%>%
  slice_sample(n = 500,
               by = cyl,
               replace = TRUE,
               weight_by = weight)

```

```{r}
mtcars %>%
  slice_max(order_by = hp, n = 50) %>%
  slice_max(order_by = drat, n = 20)
```


```{r}
?dplyr_by
```


```{r}
?dplyr_data_masking
```

data masking is a part of the tidyverse's non-standard evaluation, tidy evaluation. It allows us to refer to objects in our data without the laborious df$col_name syntax 
```{r}
mtcars$mpg

mtcars <- data("mtcars")


mtcars <- tibble::tibble(mtcars)

mtcars %>%
  summarise(mean = mean(mpg))

mtcars %>%
  mutate(mpg = "x")

mtcars$mpg <- "y"
mtcars
```

This stuff is more useful when building functions, or dealing with multiple arguments (where it becomes incredibly useful)


```{r}
mean(mtcars$mpg)

mtcars %>%
  mean(mpg)

mtcars %>%
  summarise(mean = mean(mpg))

mtcars %>%
  summarise(mean = mean(mtcars$mpg))
```

```{r}
mtcars %>%
  filter(cyl > 2, mpg > 16, disp < 250)
```

```{r}
mtcars[mtcars$mpg < 15,]
```

